---
title: "Assignment5_Data607"
author: "Sadia Perveen"
date: "9/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


install.packages("tidyr", repos = "http://cran.us.r-project.org")
install.packages("dplyr", repos = "http://cran.us.r-project.org")

require(tidyr)
require(dplyr)
```
Read the file in from github using read.csv

```{r}
csvFile <- "https://raw.githubusercontent.com/sadia-perveen/DATA607-Assignment-5/master/ArrivalDelays.csv"
aD <- readr::read_csv(csvFile)
aD
```

Clean up the data:
1. Assign proper columns names from column 1 and 2.
2. Get rid of the empty rows.
3. Populate the proper airlines in row 2 and 4.
```{r}
names(aD)[1] <- "Airline"
names(aD)[2] <- "Status"
aD
aD <- aD[rowSums(is.na(aD)) != ncol(aD), ]
aD
aD$Airline[2] <- "ALASKA"
aD$Airline[4] <- "AM WEST"
aD
```

Use the tidyr pivot_longer function to conver data from wide to long. 
The follow code will take our data aD, and convert all variables except Airline and Status to rows.

```{r}
aD <- pivot_longer(aD, -c("Airline", "Status"), names_to = "Location", values_to = "NumFlights")
aD
```

Now we can start doing some data comparison.

1. We group our data aD by Airline and Status, summarising the total number of flight, by summing the NumFlights column based on the grouping.
2. Then we spread out our data, based on Status and total flights for a more asthetic look, this will also allow for us to easily get the total delayed and total on time percentatges for our airlines.
3.We convert our delayed and on time columns as numericals to conduct arithmetic operations.
4. Finally we get the total onTimePercantage and delayedPercentage. 
5. The final result shows us total onTime and Delayed percantage for both our Airline so we can easily compare the two.
```{r}
statusTotals <- data.frame(aD %>% group_by(Airline, Status) %>% summarise(totalFlights = sum(NumFlights)))

statusTotals <- statusTotals %>% spread(Status, totalFlights)

statusTotals$delayed <- as.numeric(statusTotals$delayed)

statusTotals$`on time` <- as.numeric(statusTotals$`on time`)

statusTotals$onTimePercentage <- statusTotals$`on time`/(statusTotals$`on time` + statusTotals$delayed) * 100
statusTotals$delayedPercentage <- statusTotals$delayed/(statusTotals$`on time` + statusTotals$delayed) * 100

statusTotals
```

